---
title: "Loading and Preprocessing MRD dataset"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# libraries
library(flowCore)
library(tidyverse)
library(tidytof)
library(readxl)
library(keras)

# parameters
data_path <- here::here("data-raw", "data")
md_path <- here::here("data-raw", "Metadata.xlsx")

timepoint_levels <- c("Dx", "D8", "D15", "D33", "D78", "Rx")
risk_levels <- c("SR", "IR", "HR")
```

# Read in data

```{r}
# metadata
metadata <- 
  md_path %>% 
  read_excel() %>% 
  janitor::clean_names()

metadata
```


```{r}
# single-cell data

# function to clean channel names
clean_channels <- function(channel_name){ 
  new_name <- 
    channel_name %>% 
    str_to_lower() %>% 
    str_extract("_.+$") %>% 
    str_sub(start = 2L)
  }

all_data <- 
  data_path %>% 
  tof_read_fcs()

colnames(all_data)
```

We can see that the all data has some irrelevant channels. We can filter them out.

```{r}
all_data <- 
  all_data %>% 
  rename_with(.fn = clean_channels, .cols = contains("_")) %>% 
  rename_with(.fn = str_to_lower) %>% 
  rename(file_name = name) %>% 
  select(
    -time, 
    -length, 
    -`127i`, 
    -`89y`, 
    -dna1, 
    -dna2, 
    -width, 
    -residual, 
    -offset, 
    -beaddist,
    -center, 
    -cd16_cd33,
    -cd235_cd61, 
    -cisplatin, 
    -contains("bc")
  )

colnames(all_data)
```


# Preprocessing 

We can also preprocess the data with the asinh transformation and by removing the noise that the cytof software adds to every set of counts. 

```{r}
all_data <- 
  all_data %>%
  tof_preprocess()
```

In order to put the data through a feedforward network, we will also have to scale/center each feature (cell-wise). 

# Integrate single-cell and metadata

From here, we can annotate each cell with some variables from the metadata table `md`. 

```{r}
all_data <- 
  all_data %>% 
  left_join(y = metadata, by = "file_name")

all_data
```

We can also make one more data structure that will be useful later. 

```{r}
# create an easy lookup for which patients have which mrd risk
patient_risk <- 
  all_data %>% 
  select(patient_id, mrd_risk) %>% 
  distinct() %>% 
  drop_na()

patient_risk
```


## Leveling factors

We can also put some factors in order (like `timepoint` and `mrd_risk`) when they have a natural interpretation as "ordered."

```{r}
all_data <- 
  all_data %>% 
  mutate(
    timepoint = factor(timepoint, levels = timepoint_levels), 
    mrd_risk = factor(mrd_risk, levels = risk_levels)
  )
```


# Some quick EDA

## number of patients 

```{r}
num_patients <- 
  all_data %>% 
  count(patient_id) %>% 
  nrow()
```

We have samples from `r num_patients` unique patients. 

## Number of samples 

```{r}
num_samples <- 
  all_data %>% 
  count(file_name) %>% 
  nrow()

```

There are a total of `r num_samples` samples. This means that each patient has multiple samples taken (at different timepoints)

## Number of patients in each risk group

```{r}
all_data %>% 
  count(patient_id, mrd_risk) %>% 
  count(mrd_risk, name = "num_patients")
```

Here, we can see that one patient doesn't have an mrd risk assessment. So, we can omit the cells from this patient in the rest of our analysis. 

```{r}
all_data <- 
  all_data %>% 
  filter(mrd_risk != "n.a.")
```

## Number of patients at each timepoint

```{r}
all_data %>% 
  count(patient_id, timepoint) %>% 
  count(timepoint, name = "num_patients") %>% 
  arrange(-num_patients)
```

So we can see that only 1 patient has the Day 33 and Day 78 time points, and most have only 3 timepoints: 
- Dx
- D8
- D15

In addition, it seems that 6 patients also have samples from the relapse timepoint. 


## Number of cells per timepoint

```{r}
all_data %>% 
  mutate(timepoint = factor(timepoint, levels = timepoint_levels)) %>% 
  count(timepoint, name = "num_cells") %>% 
  ggplot(aes(x = timepoint, y = num_cells)) + 
  geom_point()
```

So we can see that we have way more cells at the first timepoint on average than at the later timepoints (with a resurgence at relapse).

## Number of cells per patient 

```{r}
patient_cells <- 
  all_data %>% 
  count(patient_id, name = "num_cells") %>% 
  left_join(patient_risk, by = "patient_id") %>% 
  mutate(patient_id = fct_reorder(patient_id, -num_cells)) %>% 
  arrange(-num_cells)

patient_cells

patient_cells %>% 
  ggplot(aes(x = patient_id, y = num_cells, color = mrd_risk)) + 
  geom_point() + 
  scale_y_log10() + 
  coord_flip()
```

So we can see that some patients have a very small number of cells (several hundred) whereas some patients have up to 500K. 

## Number of cells per patient per timepoint

```{r}
patient_timepoint_df <- 
  all_data %>% 
  count(patient_id, timepoint, name = "num_cells") %>% 
  left_join(patient_risk, by = "patient_id") %>% 
  mutate(timepoint = factor(timepoint, levels = timepoint_levels)) %>% 
  arrange(-num_cells)

patient_timepoint_df 

patient_timepoint_df %>% 
  ggplot(aes(x = timepoint, y = num_cells, color = mrd_risk)) + 
  geom_line(aes(group = patient_id), color = "gray", alpha = 0.5) + 
  geom_point()
```

It would probably be useful to look at these plots as a percentage of blasts relative to the entire sample. 

## Number of cells per risk group

```{r}
all_data %>% 
  count(mrd_risk, name = "num_cells") %>% 
  arrange(-num_cells)
```

## Number of cells per risk group per timepoint

```{r}
risk_timepoint <- 
  all_data %>% 
  count(mrd_risk, timepoint, name = "num_cells") %>% 
  arrange(timepoint, -num_cells)

risk_timepoint 

risk_timepoint %>% 
  filter(timepoint %in% c("Dx", "D8", "D15", "Rx")) %>% 
  ggplot(aes(x = timepoint, y = num_cells, color = mrd_risk)) + 
  geom_line(aes(group = mrd_risk)) + 
  geom_point() + 
  scale_y_log10()
```


We can see that the high-risk patients tend to have the highest number of blasts, which makes sense. And from the plot immediately above, we can also see that high-risk and standard-risk groups tend to have the highest number of cells especially at the D8 and D15 timepoints, and that HR is higher at relapse than IR. 

# Save the data

Then we can save the pre-processed data for use later. 

```{r}
all_data %>% 
  write_rds(file = here::here("data", "all_data.rds"))
```






